### 2025-07-15 sections/1-introduction.tex
down to 10 pages

Examples of such datasets include genomic databases, time-series signals, and neural network embeddings.
Large language models such as GPT~\cite{2020arXiv200514165B, OpenAI2023GPT4TR} and LLAMA-2~\cite{Touvron2023Llama2O}, and image embedding models~\cite{radford2021learning, dosovitskiy2020image} are a common source of neural network embeddings.
Among biological datasets,  SILVA 18S~\cite{10.1093/nar/gks1219} includes ribosomal DNA sequences from approximately 2.25 million genomes, reaching 50,000 letters when aligned.
Among time-series datasets, the RadioML dataset~\cite{oshea2018radioml} contains approximately 2.55 million samples of synthetically-generated radio frequency signals.

@@

Given some measure of similarity between data points, two standard similarity search paradigms are $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN).
The former retrieves the $k$ most similar points to a query, while the latter retrieves all points within a similarity threshold $\rho$ of a query.
While prior work has used the term \textit{approximate} search to refer to $\rho$-NN, we use it more narrowly: an {approximate} search algorithm is any algorithm that yields less than perfect recall, whereas an \textit{exact} algorithm guarantees perfect recall.

$k$-NN search is one of the most widely-used methods in classification and recommendation~\cite{fix1952discriminatory, cover1967nearest}, but na\"{i}ve implementations are prohibitively slow because their time complexity is linear in the dataset's cardinality.
Existing fast algorithms are often approximate~\cite{gao2023high}, which may suffice for some applications, but the need for efficient and \textit{exact} search remains~\cite{ukey2023survey}.
For example, for a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but may be sensitive to local perturbations for smaller values of $k$.

For example, approximate methods may fail in majority-voting classifiers with small $k$ or poorly separated classes~\cite{zhang2022imbalanced}, and distance functions like cosine similarity, which violate the triangle inequality, have shown poor performance in biomedical applications~\cite{hu2016distance}.

@@

Hierarchical Navigable Small World networks~\cite{malkov2016hnsw} rely on navigable small world (NSW) networksand skip lists.
HNSW builds a multi-layered small-world graph~\cite{kleinberg2000navigation, boguna2009navigability} of the dataset, with queries and elements placed at each layer according to an exponentially-decaying probability distribution.
Each node connects to its $M$ nearest neighbors (a tunable parameter) in the in the graph, where $M$. Search proceeds top-down, using a greedy search at each layer. Accuracy is controlled by the $efSearch$ hyperparameter, which specifies how many candidates are explored during search.

InVerted File indexing (IVF)~\cite{faissivf, sacks1987multikey, kent1990signature} partitions data into high-dimensional Voronoi cells, and searches exhaustively only within cells near the query.
The number of clusters is set via the $n_{list}$ parameter; accuracy is improved by probing additional neighboring cells, controlled by the $n_{probe}$ parameter. This approach resembles earlier entropy-scaling methods~\cite{yu2015entropy}.

ANNOY~\cite{annoy} uses random projection and tree building. 
At each intermediate node of the tree, a random hyperplane, defined by two sampled points, splits the space.  Multiple such trees form a forest; increasing the number of trees improves recall but increases search time.

>>>>

Examples of such datasets include genomic databases, time-series signals, and neural network embeddings such as GPT~\cite{2020arXiv200514165B, OpenAI2023GPT4TR}, LLAMA-2~\cite{Touvron2023Llama2O}, and image embedding models~\cite{radford2021learning, dosovitskiy2020image}.
Among biological datasets,  SILVA 18S~\cite{10.1093/nar/gks1219} includes ribosomal RNA sequences from approximately 2.25 million genomes, reaching 50,000 letters when aligned.
Among time-series datasets, the RadioML dataset~\cite{oshea2018radioml} contains approximately 2.55 million samples of synthetically-generated radio frequency signals.

@@

Given some measure of similarity between data points, two standard similarity search paradigms are $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN).
The former retrieves the $k$ points closest to a query, while the latter retrieves all points within a similarity threshold $\rho$ of a query.
While prior work has used the term \textit{approximate} search to refer to $\rho$-NN, we use it more narrowly: an {approximate} search algorithm is any algorithm that yields imperfect recall, whereas an \textit{exact} algorithm guarantees perfect recall.

$k$-NN search is one of the most widely-used methods in classification and recommendation~\cite{fix1952discriminatory, cover1967nearest}, but na\"{i}ve implementations are prohibitively slow because due to linear time complexity.
Existing fast algorithms are often approximate~\cite{gao2023high}, which may suffice for some applications, but the need for efficient and \textit{exact} search remains~\cite{ukey2023survey}.
In a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but be sensitive to local perturbations for smaller values of $k$.

@@

Hierarchical Navigable Small World networks~\cite{malkov2016hnsw} rely on navigable small world (NSW) networks and skip lists. InVerted File indexing (IVF)~\cite{faissivf, sacks1987multikey, kent1990signature} partitions data into high-dimensional Voronoi cells, and searches exhaustively only within cells near the query. ANNOY~\cite{annoy} uses random projection and tree building.
At each intermediate node of the tree, a random hyperplane, defined by two sampled points, splits the space.  Multiple such trees form a forest; increasing the number of trees improves recall but increases search time.
